Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-09 ind_files/02_BD_processes.tex
Fix in-text citation spacing and paragraph spacing

Here, I provide a brief, informal introduction to basic notions in stochastic processes that will be helpful for some technical portions of this thesis. Interested readers looking for a more comprehensive source can refer to standard mathematics texts such as \cite{oksendal_stochastic_1998} or \cite{karatzas_brownian_1998} for a more rigorous treatment of the mathematical foundations, or physics-style texts such as \cite{gardiner_stochastic_2009} or \cite{van_kampen_stochastic_1981} for useful tools and techniques to study real systems.

@@

\subsection{SDEs and the Fokker-Planck equation}\label{intro_SDE}
Stochastic systems which change continuously (in the state space) can be described in terms of a `stochastic differential equation' (SDE), which here is interchangeable with the phrase `It\^o process'. An SDE for a stochastic process $\{X_t\}_{t \geq 0}$ is an equation of the form
\label{ito_SDE_integral}
X_t = \int\limits_{0}^{t} F(s,X_s)ds + \int\limits_{0}^{t} G(s,X_s)dB_s
where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration and for our cases will usually be the Euclidean norm. For biological systems, both of these conditions will usually be satisfied, and so we assume going further that all our SDEs are always well-defined and admit solutions.} In the physics literature, $F$ and $G$ are often called the `drift' and `diffusion' of the process respectively. However, we will not use this terminology here due to  potential confusion with genetic drift (which actually corresponds to the `diffusion' in the physics terminology). $B_t$ is the so-called `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{B_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $B_t$ as recording the position of a small pollen grain at time $t$, then $B_t$ can be formally thought of as a process that has the following properties:
\begin{itemize}
	\item It starts at the origin, \emph{i.e} $B_0 = 0$. This is a harmless assumption made for convenience and amounts to a choice of coordinate system.
	\item It moves continuously, without sudden jumps across regions of space, \emph{i.e} the map $t \to B_t$ is continuous. This simply says that our pollen grain moves short distances in short intervals of time.
	\item The future movement is independent of past history. That is, given times $u < s < t$, the displacement $B_t - B_s$ is independent of the past position $B_u$.
	\item The movement is directionless and random, and displacement is normally distributed. More precisely, given two times $s < t$, the displacement $B_t - B_s$ follows a normal distribution with a mean of $0$ (this is the `directionless' part) and a variance of $t-s$ (this is the `random' part).
\end{itemize}
It can then be shown that since the motion is equally likely to be in any direction, the expected position at any point of time is the same as the initial position, \emph{i.e} $\mathbb{E}[B_t | B_0] = B_0 = 0$.\\
The second integral in equation \eqref{ito_SDE_integral} is It\^o's `stochastic integral', and is to be interpreted in the following sense: Fix a time $T > 0$. Partition the interval $[0,T]$ into $n$ intervals of the form $[t_i,t_{i+1}]$ such that $0 = t_0 < t_1 < t_2 < \ldots < t_n = T$. Then, the (It\^o) stochastic integral from $0$ to $T$ can be thought of as:
\int\limits_{0}^{T} G(s,X_s)dB_s \coloneqq \lim_{n \to \infty} \sum\limits_{i=1}^{n}G(t_i,X_{t_i})(B_{t_{i+1}}-B_{t_i})  
That is to say, it is obtained by making successively finer partitions of the form $[t_i,t_{i+1}]$, and then computing the `area of the rectangle' formed with $B_{t_{i+1}}-B_{t_i}$ and $G(t_i,X_{t_i})$ as sides. This should look similar to the classic Riemann integral, with the uniform width $t_{i+1}-t_i$ of the Riemann integral replaced by a random width corresponding to the random displacement of a Brownian particle during the uniform time interval $[t_i,t_{i+1}]$.\\

@@

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature \citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics \citep{ewens_mathematical_2004} and pure mathematics \citep{oksendal_stochastic_1998} literature. If the function $G$ is independent of $x$, then it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

\subsection{Density-dependence and the intuition for system-size expansions in ecology}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations in ecology, at an individual level, births and deaths of individuals are affected by local population density and not directly by the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ by the introduction of a `system size parameter' $K$. In ecology, this parameter will be some fundamental limit on resources, such as habitat size or carrying capacity. In physics and chemistry, it is usually the total volume of a container in which physical or chemical reactions take place. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms. A similar approximation is well-known (ever since Fisher) in theoretical population genetics, where it goes by the name of the `diffusion approximation'\citep{ewens_mathematical_2004} or `continuum limit' \citep{czuppon_understanding_2021}, and has been heavily used by Kimura \citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed \citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter (\emph{i.e.} is closer to a Kramers-Moyal expansion than a Van Kampen expansion).

@@

\section{Warm up: One-dimensional processes for population size}\label{sec_1D_processes}
The simplest birth-death processes are those in which the state at any time can be characterized by a single number. Populations of identical individuals are an obvious example of such a system. I will use this toy system as an illustration of the techniques that will be used for the actual problems we intend to tackle in the next sections. The mathematics below are adapted from sections 6.3 and 7.2 of \citep{gardiner_stochastic_2009}.

@@

Consider a population of identical individuals subject to some ecological rules that affect individuals' birth and death rates. Since all individuals are identical, we can only really track the population size through time. The population as a whole at any time $t$ can thus be characterized by a single number - its population size (Figure \ref{fig_1D_pop_description}). Imagine further that if a population has $n$ identical individuals, then, from the ecological rules, we can determine a \emph{birth rate} $b(n)$, which gives us a measure of the probability that a new individual will be born and the population size becomes $n+1$ `per unit time'. One must be slightly precise about what exactly they mean when they say `per unit time' since there are no discrete `time steps' for individuals to be born. Here, by `birth rate', we mean the probability that there will be a birth (and no death) per an \emph{infinitesimal} amount of time. More formally, letting $N_t$ denote the random variable representing the population size at time $t$ and letting $\mathbb{P}(E)$ denote the probability (in the common-sense usage) of an event $E$, the birth rate $b(n)$ of a population with population size $n$ is the quantity
\begin{equation}
\label{1D_birthrate_defn}
b(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n+1 | N_{t}=n\right)
\end{equation}
\\
Exactly analogously, we also assume we can define a \emph{death rate} $d(n)$ of a population of $n$ individuals as the quantity
\begin{equation}
\label{1D_deathrate_defn}
d(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n-1 | N_{t}=n\right)
\end{equation}
An alternative, perhaps more intuitive characterization, of these same quantities is the following: If we have a population of size $n$, and we know that \emph{either a birth or a death} has just occurred, then, the probability that the event that occurred is a birth is
\begin{equation*}
    \mathbb{P[\textrm{ birth } | \textrm{ something happened }]} = \frac{b(n)}{b(n)+d(n)}
\end{equation*}
and the probability that the event was instead a death is given by
\begin{equation*}
    \mathbb{P[\textrm{ death } | \textrm{ something happened }]} = \frac{d(n)}{b(n)+d(n)}
\end{equation*}
\\
\begin{example}\label{ex_1D_stoch_logistic}
Consider the case where the per-capita birth rate is a constant $\lambda > 0$, \emph{i.e}, $b(n) = \lambda n$, and the per-capita death rate has the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Taking the difference between the birth and death rates, we obtain $b(n) - d(n) = (\lambda - \mu)n\left(1-\frac{n}{K}\right)$, where, identifying $r=\lambda-\mu$, we obtain the familiar logistic equation on the RHS. Note, however, that the population itself is stochastic, whereas the logistic equation is a deterministic description.
\end{example}
Now, let $P(n,t)$ be the probability that the population size is $n$ at time $t$. We wish to have an equation to describe how $P(n,t)$ changes with time - this will provide a probabilistic description of how we expect the population size to change over time.
\myfig{0.5}{Media/3.1_BD_process_1D.png}{\textbf{Schematic description of a one-dimensional birth-death process}. Consider a population of identical individuals. The state of the system can be described by a single number, in this case, the population size. Births and deaths result in changes in the total population size, and the birth and death rates are dependent on the current population size.}{fig_1D_pop_description}

@@

\subsection{Stochastic fluctuations and the weak noise approximation}\label{sec_1D_WNA}
% The mathematics in this section is taken from \cite{gardiner_stochastic_2009} with minor modifications\footnote{Gardiner uses a parameter $\epsilon>0$ such that noise increases with an increase in $\epsilon$. I have used a parameter $K>0$ such that noise decreases with an increase in $K$, which is more biologically relevant. Gardiner's exact results can be recovered by substituting $\epsilon = \frac{1}{\sqrt{K}}$}.
If we assume the noise is \emph{weak}, then we can go still further with analytic techniques by measuring fluctuations from the deterministic expectations, albeit with some slightly cumbersome calculations to arrive at the final expressions. We will grit our teeth and get through the algebra below, with my promise that the final answer is neat and easy to handle. It is clear that as $K \to \infty$, equation \eqref{1D_SDE} describes a deterministic process, obtained as the solution to
\label{1D_det_limit}
    \frac{dx}{dt} = A^{-}(x) = b_{K}(x) - d_{K}(x)
This is a very intuitive equation, saying that the rate of change of the population is equal to the birth rate minus the death rate. Let the solution of this equation  be given by $\alpha(t)$, so that $\frac{d{\alpha}}{dt}(t) = A^{-}(\alpha(t))$.\\
We can now measure (scaled) fluctuations from the deterministic solution $\alpha$ through a new variable $y=\sqrt{K}\left(x-\alpha(t)\right)$. For notational clarity, we will also introduce a new time variable $s=t$ which is equal to the original time variable (this is just so the equations look clearer). Let the probability density function of this new variable be given by $\Tilde{P}(y,s)$. In summary, we have introduced the variables:
    y &= \sqrt{K}\left(x-\alpha(t)\right)\\
    s &= t\\
    \Tilde{P}(y,s) &= \frac{1}{\sqrt{K}}P(x,t)
Note that by ordinary rules of variable substitution, we have:
    \frac{\partial \Tilde{P}}{\partial t} &= \frac{\partial \Tilde{P}}{\partial y}\frac{\partial y}{\partial t} + \frac{\partial \Tilde{P}}{\partial s}\frac{\partial s}{\partial t}\nonumber\\
    &=\frac{\partial \Tilde{P}}{\partial y}\left( -\sqrt{K}\frac{d\alpha}{dt}\right) + \frac{\partial \Tilde{P}}{\partial s}\nonumber\\
    &= -\sqrt{K}A^{-}(\alpha(s))\frac{\partial \Tilde{P}}{\partial y} + \frac{\partial \Tilde{P}}{\partial s}\label{weak_noise_expansion_first_term}
and
\label{weak_noise_expansion_second_term}
    \frac{\partial }{\partial y} = \frac{1}{\sqrt{K}}\frac{\partial }{\partial x}
Reformulating \eqref{1D_FPE} in terms of $y,s$ and $\Tilde{P}$ and substituting \eqref{weak_noise_expansion_first_term} and \eqref{weak_noise_expansion_second_term} yields:
-A^-(\alpha)\frac{\partial\Tilde{P}}{\partial x} + \frac{\partial \Tilde{P}}{\partial s} &= -\sqrt{K}\frac{\partial}{\partial y}\left(A^-(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)+\frac{1}{2}\frac{\partial^2}{\partial y^2}\left(A^+(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)\nonumber\\
\Rightarrow \frac{\partial \Tilde{P}}{\partial s} &= -\frac{\partial}{\partial y}\left[\sqrt{K}\left(A^-(\alpha+\frac{y}{\sqrt{K}})-A^-(\alpha)\right)\Tilde{P}\right]+\frac{1}{2}\frac{\partial^2}{\partial y^2}\left(A^+(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)\label{weak_noise_exact_equation}
We are now ready to make a weak noise `expansion'. We do so by assuming that $\Tilde{P}$, $A^-(\alpha + \frac{y}{\sqrt{K}})$, and $A^+(\alpha+\frac{y}{\sqrt{K}})$ can be approximated by series expansions in $\frac{1}{\sqrt{K}}$ of the form:
    \Tilde{P} &= \sum\limits_{n=0}^{\infty}\Tilde{P}_n\left(\frac{1}{\sqrt{K}}\right)^n\\
    A^-\left(\alpha(s) + \frac{y}{\sqrt{K}}\right) &= \sum\limits_{n=0}^{\infty}A^-_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\\
    A^+\left(\alpha(s) + \frac{y}{\sqrt{K}}\right) &= \sum\limits_{n=0}^{\infty}A^+_n(s)\left(\frac{y}{\sqrt{K}}\right)^n
with $A^-_0(s) = A^-(\alpha(s)), A^+_0(s) = A^+(\alpha(s))$. These could be Taylor expansions, for example, but the exact form of the coefficients is irrelevant as long as it is known to us, so any expansion will work. We can now substitute these series expansions into \eqref{weak_noise_exact_equation} to obtain:
\begin{split}
\label{weak_noise_full_series_expansion}
\sum\limits_{n=0}^{\infty}\left(\frac{1}{\sqrt{K}}\right)^n\frac{\partial \Tilde{P}_n}{\partial s} = 
-\frac{\partial}{\partial y}\left[\sqrt{K}\left( \sum\limits_{n=1}^{\infty}A^-_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\right)\left(\sum\limits_{m=0}^{\infty}\Tilde{P}_m\left(\frac{1}{\sqrt{K}}\right)^m\right)\right] \\ + 
\frac{1}{2}\frac{\partial^2}{\partial y^2}\left[\left(\sum\limits_{n=0}^{\infty}A^+_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\right)\left(\sum\limits_{m=0}^{\infty}\Tilde{P}_m\left(\frac{1}{\sqrt{K}}\right)^m\right)\right]
\end{split}
We can now compare the coefficients of $K^{-n/2}$ for each $n$ in order to arrive at approximations in the series expansion, the idea being that you neglect all terms which are of order greater than $\mathcal{O}(K^{-m/2})$ for some $m$ according to the desired precision.\\
We observe that for any fixed $r$, the coefficient of $K^{-r/2}$ on the LHS is $\frac{\partial \Tilde{P}_r}{\partial s}$. On the RHS, the coefficients of $K^{-r/2}$ in the second term have the form $\Tilde{P}_{m}A^+_{n}y^n$, subject to the constraint that $m+n=r$. Furthermore, all such terms (and only such terms) are coefficients of $K^{-r/2}$. Thus, after grouping, the coefficient of $K^{-r/2}$ from the second terms of the RHS of \eqref{weak_noise_full_series_expansion} is precisely

>>>>

Here, I provide a brief, informal introduction to basic notions in stochastic processes that will be helpful for some technical portions of this thesis. Interested readers looking for a more comprehensive source can refer to standard mathematics texts such as~\cite{oksendal_stochastic_1998} or~\cite{karatzas_brownian_1998} for a more rigorous treatment of the mathematical foundations, or physics-style texts such as~\cite{gardiner_stochastic_2009} or~\cite{van_kampen_stochastic_1981} for useful tools and techniques to study real systems.

@@

\subsection{SDEs and the Fokker-Planck equation}\label{intro_SDE}
Stochastic systems which change continuously (in the state space) can be described in terms of a `stochastic differential equation' (SDE), which here is interchangeable with the phrase `It\^o process'. An SDE for a stochastic process $\{X_t\}_{t \geq 0}$ is an equation of the form
\begin{equation}
\label{ito_SDE_integral}
X_t = \int\limits_{0}^{t} F(s,X_s)ds + \int\limits_{0}^{t} G(s,X_s)dB_s
\end{equation}
where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration and for our cases will usually be the Euclidean norm. For biological systems, both of these conditions will usually be satisfied, and so we assume going further that all our SDEs are always well-defined and admit solutions.} In the physics literature, $F$ and $G$ are often called the `drift' and `diffusion' of the process respectively. However, we will not use this terminology here due to  potential confusion with genetic drift (which actually corresponds to the `diffusion' in the physics terminology). $B_t$ is the so-called `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{B_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $B_t$ as recording the position of a small pollen grain at time $t$, then $B_t$ can be formally thought of as a process that has the following properties:
\begin{itemize}
	\item It starts at the origin, \emph{i.e} $B_0 = 0$. This is a harmless assumption made for convenience and amounts to a choice of coordinate system.
	\item It moves continuously, without sudden jumps across regions of space, \emph{i.e} the map $t \to B_t$ is continuous. This simply says that our pollen grain moves short distances in short intervals of time.
	\item The future movement is independent of past history. That is, given times $u < s < t$, the displacement $B_t - B_s$ is independent of the past position $B_u$.
	\item The movement is directionless and random, and displacement is normally distributed. More precisely, given two times $s < t$, the displacement $B_t - B_s$ follows a normal distribution with a mean of $0$ (this is the `directionless' part) and a variance of $t-s$ (this is the `random' part).
\end{itemize}
It can then be shown that since the motion is equally likely to be in any direction, the expected position at any point of time is the same as the initial position, \emph{i.e} $\mathbb{E}[B_t | B_0] = B_0 = 0$.

The second integral in equation \eqref{ito_SDE_integral} is It\^o's `stochastic integral', and is to be interpreted in the following sense: Fix a time $T > 0$. Partition the interval $[0,T]$ into $n$ intervals of the form $[t_i,t_{i+1}]$ such that $0 = t_0 < t_1 < t_2 < \ldots < t_n = T$. Then, the (It\^o) stochastic integral from $0$ to $T$ can be thought of as:
\begin{equation*}
\int\limits_{0}^{T} G(s,X_s)dB_s \coloneqq \lim_{n \to \infty} \sum\limits_{i=1}^{n}G(t_i,X_{t_i})(B_{t_{i+1}}-B_{t_i})  
\end{equation*}
That is to say, it is obtained by making successively finer partitions of the form $[t_i,t_{i+1}]$, and then computing the `area of the rectangle' formed with $B_{t_{i+1}}-B_{t_i}$ and $G(t_i,X_{t_i})$ as sides. This should look similar to the classic Riemann integral, with the uniform width $t_{i+1}-t_i$ of the Riemann integral replaced by a random width corresponding to the random displacement of a Brownian particle during the uniform time interval $[t_i,t_{i+1}]$.

@@

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature~\citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics~\citep{ewens_mathematical_2004} and pure mathematics~\citep{oksendal_stochastic_1998} literature. If the function $G$ is independent of $x$, then it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

\subsection{Density-dependence and the intuition for system-size expansions in ecology}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations in ecology, at an individual level, births and deaths of individuals are affected by local population density and not directly by the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ by the introduction of a `system size parameter' $K$. In ecology, this parameter will be some fundamental limit on resources, such as habitat size or carrying capacity. In physics and chemistry, it is usually the total volume of a container in which physical or chemical reactions take place. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms. A similar approximation is well-known (ever since Fisher) in theoretical population genetics, where it goes by the name of the `diffusion approximation'\citep{ewens_mathematical_2004} or `continuum limit'~\citep{czuppon_understanding_2021}, and has been heavily used by Kimura~\citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed~\citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter (\emph{i.e.} is closer to a Kramers-Moyal expansion than a Van Kampen expansion).

@@

\section{Warm up: One-dimensional processes for population size}\label{sec_1D_processes}
The simplest birth-death processes are those in which the state at any time can be characterized by a single number. Populations of identical individuals are an obvious example of such a system. I will use this toy system as an illustration of the techniques that will be used for the actual problems we intend to tackle in the next sections. The mathematics below are adapted from sections 6.3 and 7.2 of~\citep{gardiner_stochastic_2009}.

@@

Consider a population of identical individuals subject to some ecological rules that affect individuals' birth and death rates. Since all individuals are identical, we can only really track the population size through time. The population as a whole at any time $t$ can thus be characterized by a single number - its population size (Figure \ref{fig_1D_pop_description}). Imagine further that if a population has $n$ identical individuals, then, from the ecological rules, we can determine a \emph{birth rate} $b(n)$, which gives us a measure of the probability that a new individual will be born and the population size becomes $n+1$ `per unit time'. One must be slightly precise about what exactly they mean when they say `per unit time' since there are no discrete `time steps' for individuals to be born. Here, by `birth rate', we mean the probability that there will be a birth (and no death) per an \emph{infinitesimal} amount of time. More formally, letting $N_t$ denote the random variable representing the population size at time $t$ and letting $\mathbb{P}(E)$ denote the probability (in the common-sense usage) of an event $E$, the birth rate $b(n)$ of a population with population size $n$ is the quantity
\begin{equation}
\label{1D_birthrate_defn}
b(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n+1 | N_{t}=n\right)
\end{equation}
\\

Exactly analogously, we also assume we can define a \emph{death rate} $d(n)$ of a population of $n$ individuals as the quantity
\begin{equation}
\label{1D_deathrate_defn}
d(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n-1 | N_{t}=n\right)
\end{equation}
An alternative, perhaps more intuitive characterization, of these same quantities is the following: If we have a population of size $n$, and we know that \emph{either a birth or a death} has just occurred, then, the probability that the event that occurred is a birth is
\begin{equation*}
    \mathbb{P[\textrm{ birth } | \textrm{ something happened }]} = \frac{b(n)}{b(n)+d(n)}
\end{equation*}
and the probability that the event was instead a death is given by
\begin{equation*}
    \mathbb{P[\textrm{ death } | \textrm{ something happened }]} = \frac{d(n)}{b(n)+d(n)}
\end{equation*}
\\

\begin{example}\label{ex_1D_stoch_logistic}
Consider the case where the per-capita birth rate is a constant $\lambda > 0$, \emph{i.e}, $b(n) = \lambda n$, and the per-capita death rate has the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Taking the difference between the birth and death rates, we obtain $b(n) - d(n) = (\lambda - \mu)n\left(1-\frac{n}{K}\right)$, where, identifying $r=\lambda-\mu$, we obtain the familiar logistic equation on the RHS. Note, however, that the population itself is stochastic, whereas the logistic equation is a deterministic description.
\end{example}
Now, let $P(n,t)$ be the probability that the population size is $n$ at time $t$. We wish to have an equation to describe how $P(n,t)$ changes with time - this will provide a probabilistic description of how we expect the population size to change over time.
\myfig{0.5}{Media/3.1_BD_process_1D.png}{\textbf{Schematic description of a one-dimensional birth-death process}. Consider a population of identical individuals. The state of the system can be described by a single number, in this case, the population size. Births and deaths result in changes in the total population size, and the birth and death rates are dependent on the current population size.}{fig_1D_pop_description}

@@

\subsection{Stochastic fluctuations and the weak noise approximation}\label{sec_1D_WNA}
% The mathematics in this section is taken from~\cite{gardiner_stochastic_2009} with minor modifications\footnote{Gardiner uses a parameter $\epsilon>0$ such that noise increases with an increase in $\epsilon$. I have used a parameter $K>0$ such that noise decreases with an increase in $K$, which is more biologically relevant. Gardiner's exact results can be recovered by substituting $\epsilon = \frac{1}{\sqrt{K}}$}.
If we assume the noise is \emph{weak}, then we can go still further with analytic techniques by measuring fluctuations from the deterministic expectations, albeit with some slightly cumbersome calculations to arrive at the final expressions. We will grit our teeth and get through the algebra below, with my promise that the final answer is neat and easy to handle. It is clear that as $K \to \infty$, equation \eqref{1D_SDE} describes a deterministic process, obtained as the solution to
\begin{equation}
\label{1D_det_limit}
    \frac{dx}{dt} = A^{-}(x) = b_{K}(x) - d_{K}(x)
\end{equation}
This is a very intuitive equation, saying that the rate of change of the population is equal to the birth rate minus the death rate. Let the solution of this equation  be given by $\alpha(t)$, so that $\frac{d{\alpha}}{dt}(t) = A^{-}(\alpha(t))$.

We can now measure (scaled) fluctuations from the deterministic solution $\alpha$ through a new variable $y=\sqrt{K}\left(x-\alpha(t)\right)$. For notational clarity, we will also introduce a new time variable $s=t$ which is equal to the original time variable (this is just so the equations look clearer). Let the probability density function of this new variable be given by $\Tilde{P}(y,s)$. In summary, we have introduced the variables:
\begin{align*}
    y &= \sqrt{K}\left(x-\alpha(t)\right)\\
    s &= t\\
    \Tilde{P}(y,s) &= \frac{1}{\sqrt{K}}P(x,t)
\end{align*}
Note that by ordinary rules of variable substitution, we have:
\begin{align}
    \frac{\partial \Tilde{P}}{\partial t} &= \frac{\partial \Tilde{P}}{\partial y}\frac{\partial y}{\partial t} + \frac{\partial \Tilde{P}}{\partial s}\frac{\partial s}{\partial t}\nonumber\\
    &=\frac{\partial \Tilde{P}}{\partial y}\left( -\sqrt{K}\frac{d\alpha}{dt}\right) + \frac{\partial \Tilde{P}}{\partial s}\nonumber\\
    &= -\sqrt{K}A^{-}(\alpha(s))\frac{\partial \Tilde{P}}{\partial y} + \frac{\partial \Tilde{P}}{\partial s}\label{weak_noise_expansion_first_term}
\end{align}
and
\begin{equation}
\label{weak_noise_expansion_second_term}
    \frac{\partial }{\partial y} = \frac{1}{\sqrt{K}}\frac{\partial }{\partial x}
\end{equation}
Reformulating \eqref{1D_FPE} in terms of $y,s$ and $\Tilde{P}$ and substituting \eqref{weak_noise_expansion_first_term} and \eqref{weak_noise_expansion_second_term} yields:
\begin{align}
-A^-(\alpha)\frac{\partial\Tilde{P}}{\partial x} + \frac{\partial \Tilde{P}}{\partial s} &= -\sqrt{K}\frac{\partial}{\partial y}\left(A^-(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)+\frac{1}{2}\frac{\partial^2}{\partial y^2}\left(A^+(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)\nonumber\\
\Rightarrow \frac{\partial \Tilde{P}}{\partial s} &= -\frac{\partial}{\partial y}\left[\sqrt{K}\left(A^-(\alpha+\frac{y}{\sqrt{K}})-A^-(\alpha)\right)\Tilde{P}\right]+\frac{1}{2}\frac{\partial^2}{\partial y^2}\left(A^+(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)\label{weak_noise_exact_equation}
\end{align}
We are now ready to make a weak noise `expansion'. We do so by assuming that $\Tilde{P}$, $A^-(\alpha + \frac{y}{\sqrt{K}})$, and $A^+(\alpha+\frac{y}{\sqrt{K}})$ can be approximated by series expansions in $\frac{1}{\sqrt{K}}$ of the form:
\begin{align*}
    \Tilde{P} &= \sum\limits_{n=0}^{\infty}\Tilde{P}_n\left(\frac{1}{\sqrt{K}}\right)^n\\
    A^-\left(\alpha(s) + \frac{y}{\sqrt{K}}\right) &= \sum\limits_{n=0}^{\infty}A^-_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\\
    A^+\left(\alpha(s) + \frac{y}{\sqrt{K}}\right) &= \sum\limits_{n=0}^{\infty}A^+_n(s)\left(\frac{y}{\sqrt{K}}\right)^n
\end{align*}
with $A^-_0(s) = A^-(\alpha(s)), A^+_0(s) = A^+(\alpha(s))$. These could be Taylor expansions, for example, but the exact form of the coefficients is irrelevant as long as it is known to us, so any expansion will work. We can now substitute these series expansions into \eqref{weak_noise_exact_equation} to obtain:
\begin{equation}
\begin{split}
\label{weak_noise_full_series_expansion}
\sum\limits_{n=0}^{\infty}\left(\frac{1}{\sqrt{K}}\right)^n\frac{\partial \Tilde{P}_n}{\partial s} = 
-\frac{\partial}{\partial y}\left[\sqrt{K}\left( \sum\limits_{n=1}^{\infty}A^-_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\right)\left(\sum\limits_{m=0}^{\infty}\Tilde{P}_m\left(\frac{1}{\sqrt{K}}\right)^m\right)\right] \\ + 
\frac{1}{2}\frac{\partial^2}{\partial y^2}\left[\left(\sum\limits_{n=0}^{\infty}A^+_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\right)\left(\sum\limits_{m=0}^{\infty}\Tilde{P}_m\left(\frac{1}{\sqrt{K}}\right)^m\right)\right]
\end{split}
\end{equation}
We can now compare the coefficients of $K^{-n/2}$ for each $n$ in order to arrive at approximations in the series expansion, the idea being that you neglect all terms which are of order greater than $\mathcal{O}(K^{-m/2})$ for some $m$ according to the desired precision.
